Bayesian Sparsification Methods for Deep Complex-valued Networks

### 2020-06-27 paper/paper.tex
reordered the appendix

In this appendix we study the approximation proposed by \citet{molchanov_variational_2017}
for the KL divergence term \eqref{eq:improper-kl-div-real} for $\real$ Sparse Variational
Dropout. Following the logic of \citet{lapidoth_capacity_2003} we derive the expression
for $
  \tfrac{d}{d\log \alpha} K(\alpha)
$. Acknowledging that the same result was obtained by \citet[eq.~(5)]{hron_variational_2018},
we provide this appendix for the sake of completeness.
% The derivation in this section different from \citet{lapidoth_capacity_2003}.
% The results are identical: they use DCT to establish interchangeability of limits
% (sum-integral-derivative), whilst we allude to calculus theorems on differentiation
% and integration of power series within the radius of their convergence (which are
% essentially corollaries to the DCT).

@@

$. Therefore, the divergence \eqref{eq:improper-kl-div-real} has the form

@@

distributions: if $Z_{\mid J} \sim \chi^2_{m + 2J}$ for $

@@

  \mathbb{E}_{W\sim \chi^2_m(\lambda)} \log W

@@

Since $\chi^2_\nu$ is Gamma distribution $\Gamma(\tfrac\nu2, \tfrac12)$, it can be shown
that the logarithmic moment $

@@

$, where $\psi$ is the digamma function ($\psi(x) = \tfrac{d}{dx} \log \Gamma(x)$).
% \begin{equation}  \label{eq:digamma}
%   \psi(x)
%     % = \frac{d}{dx} \log \Gamma(x)
%     = \frac1{\Gamma(x)}
%       \int_0^\infty
%         u^{x-1} e^{-u} \log{u}
%       \, du
% \end{equation}
By expanding expectation of a Poisson random variable we get $
  \mathbb{E}_{W\sim \chi^2_m(\lambda)} \log W
    = \log2 + g_m\bigl(\tfrac\lambda2\bigr)
$, where

@@

    % = \mathbb{E}_{J\sim {\mathcal{P}ois}(x)}
    %   \psi\bigl(\tfrac{m + 2 J}{2}\bigr)
    = e^{- x} \sum_{j\geq 0}
        \frac{x^j}{j!}

@@

@@

Making use of the property $
  \psi(z + 1) = \psi(z) + \tfrac1z
$ of the digamma funciton for $z > 0$, we conclude that the power series in \eqref{eq:exp_digamma}
converges for any $x \geq 0$. Therefore the derivative of \eqref{eq:exp_digamma} is
given by

@@

  \frac{d}{dx} g_m(x)
    % = \mathbb{E}_{J\sim {\mathcal{P}ois}(x)}
    %     \bigl( \tfrac{J}x - 1 \bigr)
    %   \psi\bigl(\tfrac{m + 2 J}{2}\bigr)
    = - g_m(x)
      + e^{- x} \sum_{j \geq 0}
          \frac{x^j}{j!}
        \biggl(

@@

          + \frac{2}{m + 2 j}
    % = e^{- x} \sum_{j \geq 0}
    %     \frac{x^j}{j!} \frac{2}{m + 2 j}

@@

By manipulating the partial sums within \eqref{eq:g_m_series} we get

@@

  \frac{d}{dx} g_m(x)

@@

    % = e^{- x} x^{- \tfrac{m}2} \sum_{j \geq 0}
    %   \frac1{j!}
    %   \frac{x^{j + \tfrac{m}2}}{j + \tfrac{m}2}

@@

Furthermore, the functions $

@@

$ are non-decreasing on $(0, x)$ with growing $J$ and converge to $

@@

$, which implies by the Monotone Convergence Theorem that

@@

      \int_0^x t^{\bigl(\tfrac{m}2 - 1\bigr)} e^t dt

@@

  I_m

@@

  \frac{d \eqref{eq:cond_exp_expansion}}{d \lambda}
    = \frac12 \frac{d}{dx} g_m(x)

@@

    = \Bigl(\sqrt{\tfrac2\lambda}\Bigr)^m I_m\Bigl(\sqrt{\tfrac\lambda2}\Bigr)

@@

    = - \frac1{\sqrt{2\alpha}} I_1\bigl(\tfrac1{\sqrt{2\alpha}}\bigr)

@@

We compute the Monte-Carlo estimate of \eqref{eq:improper-kl-div-real} on a sample
of $10^7$ draws over an equally spaced grid of $\log \alpha$ in $[-12, +12]$ of size
$4096$. The approximation proposed by \citet{molchanov_variational_2017} is given in
\eqref{eq:improper-kl-div-real-approx}, with coefficients $
  k_1 = 0.63576
$, $
  k_2 = 1.8732
$, and $
  k_3 = 1.48695
$. The derivative of the approximation with respect to $\log \alpha$ follows
\eqref{eq:real-kl-div-deriv-log} within $4\%$ of relative tolerance, see
fig.~\ref{fig:molchanov-derivative-replica}. 
% see eq.~(14): flit overall sign, use $C = -k_1$, $1-\sigma(x) = \sigma(-x)$
\begin{equation}  \label{eq:improper-kl-div-real-approx}
  \eqref{eq:improper-kl-div-real}
% - KL(\mathcal{N}(w \mid \mu, \alpha \mu^2) \|
%     \tfrac1{\lvert w \rvert})
%   \approx
%     k_1 \sigma(k_2 + k_3 \log \alpha)
%     + C \big\vert_{C = -k1}
%     - \tfrac12 \log (1 + e^{-\log \alpha})
  % = \tfrac12 \log \alpha
  %   - \mathbb{E}_{\varepsilon \sim \mathcal{N}(1, \alpha)}
  %   \log{\lvert \varepsilon \rvert} + C
  \approx
    \frac12 \log{\bigl(1 + e^{-\log \alpha}\bigr)}
    + k_1 \sigma\bigl(- (k_2 + k_3 \log \alpha)\bigr)
    % - k_1 \bigl(1 - \sigma(k_2 + k_3 \log \alpha)\bigr)
  \,,
\end{equation}
% in parctial implementations we suggest using softplus to numerically comptue the last
% term, since it is typically implemented in a floating-point stable fashion.
Similarly, the forward difference estimate of the derivative \eqref{eq:real-kl-div-deriv-log}
very closely (up to sampling error). For sake of completeness, we compute a similar
Monte-Carlo estimate for the KL divergence term in \eqref{eq:c-vd-kl-div} for $\cplx$-valued
Variational Dropout with $\beta = 2$, fit the best approximation \eqref{eq:improper-kl-div-real-approx},
and compare it against the exact $\log \alpha$ derivative $
  \frac{d \eqref{eq:c-vd-kl-div}}{d\log \alpha}
    % = \frac{d \alpha}{d\log \alpha} \frac{d}{d\alpha} \bigl(
    %   \log \tfrac1\alpha - \mathop{Ei}\bigl(-\tfrac1\alpha\bigr)
    % = \alpha \bigl(
    %   - \tfrac1\alpha
    %   - \tfrac{e^x}{x} \bigg\vert_{x=-\tfrac1\alpha} \alpha^{-2}
    % \bigr)
    = e^{-\tfrac1\alpha} - 1
$.

@@

>>>>

In this appendix we study the approximation proposed by \citet{molchanov_variational_2017}
for the KL divergence term \eqref{eq:improper-kl-div-real}. We show empirically that
the gradient feedback provided by the approximation is sufficiently close to $
  \tfrac{d}{d\log \alpha} K(\alpha)
$, making it a viable alternative to exact gradients, computed in \citet{hron_variational_2018}.
% The derivation in this section different from \citet{lapidoth_capacity_2003}.
% The results are identical: they use DCT to establish interchangeability of limits
% (sum-integral-derivative), whilst we allude to calculus theorems on differentiation
% and integration of power series within the radius of their convergence (which are
% essentially corollaries to the DCT).

@@

$. The divergence \eqref{eq:improper-kl-div-real} has the form

@@

distributions: if $Z_{\mid J} \sim \chi^2_{k + 2J}$ for $

@@

  \mathbb{E}_{W\sim \chi^2_k(\lambda)} \log W

@@

Since $\chi^2_\nu$ is Gamma distribution $\Gamma(\tfrac\nu2, \tfrac12)$, its logarithmic
moment $

@@

$, where $\psi(x)$ is the digamma function. Hence

@@

  \mathbb{E}_{W\sim \chi^2_k(\lambda)} \log W
    = \log2 + \mathbb{E}_{J\sim {\mathcal{P}ois}(\tfrac\lambda2)}

@@

    % = \log2
    % + \sum_{j\geq 0}
    %     e^{- \tfrac{\lambda}2}
    %     \frac{\bigl(\tfrac{\lambda}2\bigr)^j}{j!}
    %   \psi\bigl(\tfrac{m+2J}{2}\bigr)

@@

We can use REINFORCE method to compute the derivative of \eqref{eq:exp_digamma} with respect
to $\lambda$ \citep{williams_simple_1992}. Since Poisson $\log$-probability is $
  \log p_\lambda(J)
    = - \tfrac{\lambda}2
      + J \log \tfrac{\lambda}2
      - \log J!
$, we get
\begin{equation}  \label{eq:exp_reinforce}
  \frac{d \eqref{eq:exp_digamma}}{d \lambda}
    =
    \mathbb{E}_{J\sim {\mathcal{P}ois}(\tfrac\lambda2)}
      \tfrac{d}{d \lambda} \log p_\lambda(J)
    % = \mathbb{E}_{J}
    %   \bigl(
    %     \tfrac2\lambda J - 1
    %   \bigr)
    %   \frac{\psi\bigl(\tfrac{m + 2 J}{2}\bigr)}2
    = \frac12 g_m\bigl(\tfrac\lambda2\bigr)
where

@@

    =
    \mathbb{E}_{J\sim {\mathcal{P}ois}(x)}
      \bigl( \tfrac{J}x - 1 \bigr)
    =
    e^{- x} \sum_{j \geq 0}
      \frac{x^j}{j!} \bigl( \tfrac{J}x - 1 \bigr)

@@

@@

By manipulating the partial sums within $g_m(x)$ and making use of the property $
  \psi(x + 1) = \psi(x) + \tfrac1x
$ of the digamma funciton for $x > 0$, we get

@@

@@

      \frac1{j!}
      \frac{x^{j + \tfrac{m}2}}{j + \tfrac{m}2}

@@

Since for any $J$ the functions $

@@

$ are positive on $(0, x)$ and upper bounded by integrable $

@@

$, which turns out to be the limit of the partial sums, the Dominated Convergence
theorem implies

@@

      \int_0^x e^t t^{(\tfrac{m}2 - 1)} dt

@@

  F_m

@@

  \frac{d \eqref{eq:exp_digamma}}{d \lambda}
    = \frac12 g_m\bigl(\tfrac\lambda2\bigr)

@@

    = \Bigl(\sqrt{\tfrac2\lambda}\Bigr)^m F\Bigl(\sqrt{\tfrac\lambda2}\Bigr)

@@

    = - \frac1{\sqrt{2\alpha}} F\bigl(\tfrac1{\sqrt{2\alpha}}\bigr)

@@

We compute the Monte-Carlo estimator of \eqref{eq:improper-kl-div-real} on a sample of $10^7$
draws over an equally spaced grid of $\log \alpha$ in $[-12, +12]$ of size $4096$. The
coefficients of the approximation, proposed by \citet{molchanov_variational_2017}, are
within $1\%$ relative tolerance of the reported therein. An equivalent expression of the
approximation is given in \eqref{eq:improper-kl-div-real-approx}.
% see eq.~(14): flit overall sign, use $C = -k_1$, $1-\sigma(x) = \sigma(-x)$
\begin{equation}  \label{eq:improper-kl-div-real-approx}
  \eqref{eq:improper-kl-div-real}
% - KL(\mathcal{N}(w \mid \mu, \alpha \mu^2) \|
%     \tfrac1{\lvert w \rvert})
%   \approx
%     k_1 \sigma(k_2 + k_3 \log \alpha)
%     + C \big\vert_{C = -k1}
%     - \tfrac12 \log (1 + e^{-\log \alpha})
  % = \tfrac12 \log \alpha
  %   - \mathbb{E}_{\varepsilon \sim \mathcal{N}(1, \alpha)}
  %   \log{\lvert \varepsilon \rvert} + C
  \approx
    \frac12 \log{\bigl(1 + e^{-\log \alpha}\bigr)}
    + k_1 \sigma\bigl(- (k_2 + k_3 \log \alpha)\bigr)
    % - k_1 \bigl(1 - \sigma(k_2 + k_3 \log \alpha)\bigr)
  \,,
\end{equation}
% in parctial implementations we suggest using softplus to numerically comptue the last
% term, since it is typically implemented in a floating-point stable fashion.
The numerically estimated derivative of the penalty term with respect to $\log \alpha$ using
forward differences seems to be following \eqref{eq:real-kl-div-deriv-log} very closely (up to
sampling error), see fig.~\ref{fig:molchanov-derivative-replica}. The derivative of
\eqref{eq:improper-kl-div-real-approx} with respect to $\log \alpha$ appears to be very
close to \eqref{eq:real-kl-div-deriv-log}. For sake of completeness, we compute a similar
Monte-Carlo estimate for $\cplx$ variational dropout KL divergence term in \eqref{eq:c-vd-kl-div}
with $\beta = 2$, fit the best approximation \eqref{eq:improper-kl-div-real-approx}, and
compare it against the exact $\log \alpha$ derivative $
  \frac{d \eqref{eq:c-vd-kl-div}}{d\log \alpha}
    % = \frac{d \alpha}{d\log \alpha} \frac{d}{d\alpha} \bigl(
    %   \log \tfrac1\alpha - \mathop{Ei}\bigl(-\tfrac1\alpha\bigr)
    % = \alpha \bigl(
    %   - \tfrac1\alpha
    %   - \tfrac{e^x}{x} \bigg\vert_{x=-\tfrac1\alpha} \alpha^{-2}
    % \bigr)
    = e^{-\tfrac1\alpha} - 1
$.

@@

\section{$\cplx$-Linear operator representation} % (fold)
\label{sec:c-linear_operator_representation}

% proof that such operators exist and are unique (well this is an obvious statement)
Consider $L \colon \cplx^m \to \cplx^n$ -- linear in $\cplx$. Then $
  L(u + jv) = L u + \iu L v
$ for any $u, v \in \real^m$, which implies that the effect of $L$ as $\cplx$-linear
operator is determined by its restriction onto $\real^n$. Let $
  F = L\vert_{\real^m}
  \colon \real^m \to \cplx^n
$ and observe that $F_r = \Re \circ F$ and $F_i = \Im \circ F$ are $\real$-linear operators
such that $F = F_r + \iu F_i$ (pointwise). Indeed,
$$
  F_r(a + \lambda b)
  % = \Re F(a + \lambda b)
  = \Re L(a + \lambda b)
  % = \Re \bigl( L a + \lambda L b \bigr)
  = \Re L a + \lambda \Re L b
  % = \Re F a + \lambda \Re F b
  = F_r a + \lambda F_r b
  \,. $$
Therefore for any $\cplx$-linear operator $L$ there are $\real$-linear operators $U, V$
such that
$$
L z 
  = (U + \iu V) (\Re z + \iu \Im z)
  % = (U + \iu V) \Re z + \iu (U + \iu V) \Im z
  % = U \Re z + \iu V \Re z + \iu \bigl(U \Im z + \iu V \Im z \bigr)
  = U \Re z - V \Im z + \iu \bigl( V \Re z + U \Im z \bigr)
  % = (U + \iu V) z
  \,. $$
Uniqueness of these operators follows, if this decomposition is applied to any $z$ with
$\Im z = 0$.

% section c-linear_operator_representation (end)
